\section{Cosine similarity as a soft relaxation of exact-match accuracy}
\label{sec:cosine}

\subsection{Embedding cosine similarity}

The \texttt{noun} field is free text (``cereal'', ``sippy cup'',
``baked beans'', \dots), so exact-string accuracy penalizes correct-but-
differently-worded answers as harshly as wrong ones. Let
$e:\mathcal{W}\to\mathbb{R}^d$ be a sentence-embedding map (MiniLM,
$d=384$) with $\lVert e(w)\rVert_2 = 1$ for all $w$ (embeddings are
L2-normalized before comparison). For ground-truth noun $g_i$ and
predicted noun $\hat{g}_i$ on clip $i$,
\[
  \mathrm{sim}(g_i,\hat{g}_i) \;=\; \cos\theta_i \;=\;
  \frac{e(g_i)\cdot e(\hat{g}_i)}{\lVert e(g_i)\rVert\,\lVert e(\hat{g}_i)\rVert}
  \;=\; e(g_i)\cdot e(\hat{g}_i) \in [-1,1],
\]
the last equality using unit normalization. Because $e$ is trained so that
semantically related words are embedded nearby, $\mathrm{sim}$ acts as a
continuous relaxation of the indicator $\mathbb{1}[g_i = \hat{g}_i]$:
identical strings give $\mathrm{sim}=1$ exactly, while near-synonyms
(``sippy cup'' vs.\ ``bottle'') give intermediate values and unrelated
words give values near $0$ or below.

\subsection{Soft accuracy and its relation to exact-match accuracy}

\begin{definition}[Soft accuracy at threshold $\tau$]
\[
  \mathrm{Acc}_{\mathrm{soft}}(\tau) \;=\;
  \frac{1}{n}\sum_{i=1}^n \mathbb{1}\!\left[\mathrm{sim}(g_i,\hat{g}_i) \ge \tau\right].
\]
\end{definition}
Exact-match accuracy is the degenerate case $\mathrm{Acc}_{\mathrm{exact}}
= \mathrm{Acc}_{\mathrm{soft}}(1)$ under an embedding where distinct
strings almost never map to identical vectors, and the mean similarity
$\overline{\mathrm{sim}} = \frac{1}{n}\sum_i \mathrm{sim}(g_i,\hat{g}_i)$
reported throughout the suite is not a threshold accuracy but a
continuous summary of the same quantity.

\begin{proposition}[Decomposition of mean similarity by exact-match status]
\label{prop:decomp}
Let $\mathcal{M} = \{i : g_i = \hat{g}_i\}$ (exact matches, contributing
$\mathrm{sim}_i = 1$) and $\mathcal{M}^c$ its complement, with
$m = |\mathcal{M}^c|/n$ the mean similarity restricted to mismatches. Then
\[
  \overline{\mathrm{sim}}
  = \mathrm{Acc}_{\mathrm{exact}} \cdot 1 \;+\; (1-\mathrm{Acc}_{\mathrm{exact}})\cdot \bar{m},
  \qquad
  \bar{m} \;=\; \frac{\overline{\mathrm{sim}} - \mathrm{Acc}_{\mathrm{exact}}}{1 - \mathrm{Acc}_{\mathrm{exact}}}.
\]
\end{proposition}
\begin{proof}
Split the sum defining $\overline{\mathrm{sim}}$ over $\mathcal{M}$ and
$\mathcal{M}^c$; every term in $\mathcal{M}$ equals $1$ by definition of
exact match, and $|\mathcal{M}|/n = \mathrm{Acc}_{\mathrm{exact}}$.
Rearranging gives $\bar m$.
\end{proof}

\subsection{Worked example (Benchmark 1)}

Benchmark~1 reports, on the annotator-5 eval set restricted to human
eating/possibly-eating clips, $\mathrm{Acc}_{\mathrm{exact}} = 70.8\%$
($n=48$) and $\overline{\mathrm{sim}} = 82.0\%$ ($n=46$; two additional
clips were excluded where the human noun was blank, so the identity below
is approximate rather than exact over an identical index set). Applying
Proposition~\ref{prop:decomp} with these figures,
\[
  \bar m \;\approx\; \frac{0.820 - 0.708}{1 - 0.708} \;\approx\; 0.384,
\]
i.e.\ on the $\approx 29\%$ of clips where Qwen's noun did \emph{not}
exactly match the human's, the two words were on average only weakly
related (cosine $\approx 0.38$) rather than close synonyms. Cross-checking
against the raw pairs reported in the suite (e.g.\ \texttt{cereal
$\to$ (none)} at $0.13$, \texttt{banana $\to$ bottle} at $0.42$,
\texttt{beans $\to$ baked beans} at $0.81$) confirms the mismatches are a
mix of outright misses (Qwen abstaining or picking an unrelated object)
and near-misses (a related but non-identical object), consistent with
$\bar m$ sitting well below $1$ but above $0$.

\begin{remark}[Why $\overline{\mathrm{sim}} \ge \mathrm{Acc}_{\mathrm{exact}}$
is not automatic]
Proposition~\ref{prop:decomp} shows $\overline{\mathrm{sim}} >
\mathrm{Acc}_{\mathrm{exact}}$ exactly when $\bar m > 0$, i.e.\ when the
model's \emph{mistakes} are, on average, still semantically closer to the
truth than two independently drawn nouns would be. This is an empirical
property of the model's error distribution, not a mathematical
guarantee --- a model that fails by naming a semantically unrelated object
every time it is wrong would have $\bar m \le 0$ and could show
$\overline{\mathrm{sim}} < \mathrm{Acc}_{\mathrm{exact}}$.
\end{remark}
